Constructing a universe for the setoid model

### 2020-07-06 paper.tex
small fixes

We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W
: (S : \mType) \to (S \to \mType) \to \mType$ and $W_p : (S : \mType) \to (S \to
\mType) \to \mProp$.
%
From $W_p$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

@@

By \emph{setoid model} we intend a class of models of type theory where
contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
relation, and dependent types are interpreted as dependent/indexed setoids.
%
A setoid model was first given for intensional type theory by M. Hofmann in
\cite{hofmann}, in order to give a semantics to extensionality principles such
as function and propositional extensionality.

@@

The setoid model thus defined validates function extensionality, a universe of
propositions with propositional extensionality, and quotient types. Therefore,
it provides a way to bootstrap and ``explain'' extensionality, since the model
construction effectively gives an implementation of various extensionality
principles it in terms of a small, completely intensional theory.

@@

In our presentation of the setoid model, contexts are given by setoids, that is
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.

@@

This definition of types in the setoid model is different from the one in
\cite{setoid99}, but it is equivalent to it \todo{add citation}. The main
difference here is in the use of a heterogeneous equivalence relation $A^\sim$
in the definition of types.

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
~
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \\
    & \quad \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
    (p : \GG^\sim\ \gamma_0\ \gamma_1) \\
    & \quad \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\
~
\\
We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), including function extensionality and a universe of strict
propositions satisfying propositional extensionality \cite{setoid99}. Note that
we don't need neither identity types nor inductive types (W-types) to do this
much.

>>>>

We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W
: (S : \mType) \to (S \to \mType) \to \mType$ and $W_p : (S : \mType) \to (S \to
\mType) \to \mProp$. The elimination principle of $W_p$ only allows defining functions into $\mProp$s.
%
From $W_p$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

@@

By \emph{setoid model} we mean a class of models of type theory where
contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
relation, and dependent types are interpreted as dependent/indexed setoids.
%
A setoid model was first given for intensional type theory by M.\ Hofmann
\cite{hofmann}, in order to provide a semantics for extensionality principles such
as function and propositional extensionality.

@@

The setoid model thus defined validates function extensionality, a universe of
propositions with propositional extensionality, and quotient types. Therefore,
it provides a way to bootstrap and ``explain'' extensionality, since the model
construction effectively gives an implementation of various extensionality
principles in terms of a small, completely intensional theory.

@@

In our presentation of the setoid model, contexts are given by setoids, that is,
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.

@@

This definition of types in the setoid model is different from the one in
\cite{setoid99}, but it is equivalent to it \cite[Section 1.6.1]{simonthesis}. The main
difference here is in the use of a heterogeneous equivalence relation $A^\sim$
in the definition of types.

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
~
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \\
    & \quad \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
    (p : \GG^\sim\ \gamma_0\ \gamma_1) \\
    & \quad \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\
~
\\
We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), including function extensionality and a universe of strict
propositions satisfying propositional extensionality \cite{setoid99}. Note that
we don't need identity types or inductive types (W-types) for this.
